---
title: "Generating missing values for simulation purposes: A multivariate amputation approach"
author: |
  | Rianne Schouten
institute: |
  | 1. University Utrecht, Department of Methodology and Statistics
  | 2. DPA Professionals, Data Science Excellence Program
date: "July 7, 2017"
output:
  beamer_presentation: 
    includes:
      in_header: preamble.tex
    theme: default
    colortheme: "beaver"

---

# Amputation

```{r, include = FALSE}
require(mice)
require(lattice)
require(MASS)
require(Matrix)
```

Amputation is the generation of missing values in complete data 

Overview:

- Why? 
- What? 
- How? 

```{r, eval = FALSE}
require(mice)
?ampute
```

# Amputation: Why? 

Evaluation of missing data methodologies:
\begin{enumerate}
\item Simulate complete data set
\item Generate missing values
\item Deal with missing data with new method
\item Compare statistical inferences 
\end{enumerate}

But also:
\begin{itemize}
\item Planned missing data survey designs
\item Investigate measurement errors
\item See effect missing data on your analyses
\end{itemize}

```{r, include = FALSE}
sigma <- nearPD(matrix(c(1, 0.5, 0.7, 
                         0.5, 1, 0.2,
                         0.7, 0.2, 1), 3, 3))$mat
sigma

set.seed(1234)
data <- mvrnorm(mu = c(0, 0, 0), 
                Sigma = sigma, 
                n = 100000)
data <- as.data.frame(data)
names(data) <- c("Income", "Minutes", "Age")
amp <- ampute(data, patterns =  c(0, 1, 1), prop = 0.5)
customer_data <- amp$amp
```

# Amputation: What? 

- Proportion
- Amputed variables

# Amputation: What? 

- Proportion
- Amputed variables
- Mechanism

\vspace{-1cm}

\begin{columns}
\begin{column}{0.5\textwidth}
\begin{align*}
\text{MCAR} &: \text{Mis not related to X or Y at all} \\
\text{MAR} &: \text{Mis related to X but not to Y} \\
\text{MNAR} &: \text{Mis related to Y}\\
\end{align*}
\end{column}

\begin{column}{0.48\textwidth}
\begin{figure}
  \includegraphics[height=60mm]{SUA.pdf}
\end{figure}
\end{column}
\end{columns}

# Amputation: What? 

```{r}
# Customers Phone Company
head(customer_data)
```

# Amputation: What? 

\begin{figure}
\centering
  \includegraphics[width=110mm]{hists_2.pdf}
\end{figure}

\vspace{-1cm}

\begin{align*}
\text{MCAR} &: Pr(Income = missing) = 0.5 \\
\text{MAR} &: Pr(Income = missing) = Age \\
\text{MNAR} &: Pr(Income = missing) = Income \\
\end{align*} 

# Amputation: What? 

- Proportion
- Amputed variables
- Mechanism
- Influencing variables
- Severity
- Missingness distribution

# Amputation: What? 

\begin{figure}
  \includegraphics[width=100mm, height=60mm]{types.pdf}
\end{figure}

# Amputation: How? 

\begin{columns}

\begin{column}{0.3\textwidth}
\begin{enumerate}
\item $Y_1$
\end{enumerate}
\end{column}

\begin{column}{0.48\textwidth}
\begin{figure}
  \includegraphics[height=60mm]{SUA_1.pdf}
\end{figure}
\end{column}

\end{columns}

# Amputation: How? 

\begin{columns}

\begin{column}{0.3\textwidth}
\begin{enumerate}
\item $Y_1$
\item $Y_2$
\end{enumerate}
\end{column}

\begin{column}{0.48\textwidth}
\begin{figure}
  \includegraphics[height=60mm]{SUA_2.pdf}
\end{figure}
\end{column}

\end{columns}

# Amputation: How? 

\begin{columns}

\begin{column}{0.3\textwidth}
\begin{enumerate}
\item $Y_1$
\item $Y_2$
\item $\dots$
\item $Y_l$
\end{enumerate}
\end{column}

\begin{column}{0.48\textwidth}
\begin{figure}
  \includegraphics[height=60mm]{SUA.pdf}
\end{figure}
\end{column}

\end{columns}

# Amputation with `ampute` 

Multivariate Amputation:
\vspace{-5mm}
\begin{figure}
  \includegraphics[height=60mm]{Scheme_large.pdf}
\end{figure}

# Amputation with `ampute` 

`
ampute(data, prop = 0.5, patterns = NULL, freq = NULL, mech = "MAR", weights = NULL, cont = TRUE, type = NULL, odds = NULL, bycases = TRUE, run = TRUE)
`

```{r}
amp <- ampute(data)
class(amp)
```

# Amputation with `ampute` 

```{r}
head(amp$amp)
```

# Documentation

```{r, eval = FALSE}
require(mice)
?ampute
```
\small

**Vignette:**

https://github.com/RianneSchouten/Amputation_with_Ampute

\vspace{5mm}
**Contact:**

Rianne Schouten, r.m.schouten\@uu.nl, rianne.schouten\@dpa.nl

\begin{figure}
\includegraphics[width=3cm]{uu_logo}\hspace*{2mm}
\includegraphics[width=3cm]{dpa_logo}
\end{figure}

# Additional slides

A missing data pattern is a specific combination of variables with missing values and variables without missing values. 

`0`: incomplete variable

`1`: complete variable

```{r}
mypat <- matrix(c(0, 0, 1,
                  0, 1, 1), 
                nrow = 2, byrow = TRUE)
```

# Additional slides

\begin{figure}
  \includegraphics[width=100mm, height=60mm]{types_shifted.pdf}
\end{figure}

# Additional slides

\begin{figure}
  \includegraphics[height=25mm]{Table.pdf}
\end{figure}
