---
title: Generating missing values with `ampute`
subtitle: And why on earth you would do that
author: |
  | Rianne Schouten
institute: |
  | 1. PhD Candidate, Utrecht University
  | 2. Developer Data & Analytics, Samen Veilig Midden-Nederland 
date: "September 1, 2018"
fontsize: 10pt
output:
  beamer_presentation: default
  theme: "default"
  colortheme: "beaver"
  toc: true
header-includes:
  - \usepackage{caption}
  - \usepackage{amsmath}
  - \usepackage{blkarray}
---

# Why on earth would you ampute?

For simulation purposes: 

1. Generate complete data (or use real dataset)
2. Generate missing values in complete data
3. Apply missing data method
4. Perform analysis and compare with complete data

But also for:

- Planned missing data survey designs
- Investigating measurement errors
- Reproducing your missing data situation

# Missing data in your dataset 

```{r, include = FALSE}
require(mice)
require(lattice)
require(MASS)
require(Matrix)

sigma <- nearPD(matrix(c(1, 0.5, 0.3, 
                         0.5, 1, 0.5,
                         0.3, 0.5, 1), 3, 3))$mat
set.seed(1234)
data <- mvrnorm(mu = c(0, 0, 0), 
                Sigma = sigma, 
                n = 1000)
data <- as.data.frame(data)
names(data) <- c("Income", "WorkingYears", "Age")
amp <- ampute(data, patterns =  matrix(c(1, 0, 1, 
                                         0, 0, 1), nrow = 2, byrow = TRUE), 
              prop = 0.8, mech = "MAR")
inc_data <- amp$amp

#data <- mvrnorm(mu = c(0, 0), 
#                Sigma = matrix(c(1, 0.7, 0.7, 1), 2, 2), 
#                n = 1000)
#data <- as.data.frame(data)
#names(data) <- c("income", "age")
#inc_data <- ampute(data, pat = c(0, 1), prop = 0.7, mech = "MCAR")$amp
```

```{r}
head(inc_data)
require(mice)
md.pattern(inc_data)
```

# Generation of missing values

\begin{columns}

\begin{column}{0.3\textwidth}
\begin{enumerate}
\item $Y_1$
\end{enumerate}
\end{column}

\begin{column}{0.48\textwidth}
\begin{figure}
  \includegraphics[height=60mm]{SUA_1.pdf}
\end{figure}
\end{column}

\end{columns}

# Generation of missing values 

\begin{columns}

\begin{column}{0.3\textwidth}
\begin{enumerate}
\item $Y_1$
\item $Y_2$
\end{enumerate}
\end{column}

\begin{column}{0.48\textwidth}
\begin{figure}
  \includegraphics[height=60mm]{SUA_2.pdf}
\end{figure}
\end{column}

\end{columns}

# Generation of missing values 

\begin{columns}

\begin{column}{0.3\textwidth}
\begin{enumerate}
\item $Y_1$
\item $Y_2$
\item $\dots$
\item $Y_l$
\end{enumerate}
\end{column}

\begin{column}{0.48\textwidth}
\begin{figure}
  \includegraphics[height=60mm]{SUA.pdf}
\end{figure}
\end{column}

\end{columns}

# Multivariate amputation with `ampute` 

\begin{figure}
  \includegraphics[height=60mm]{Scheme_large.pdf}
\end{figure}

# Multivariate Amputation with `ampute` 

Explanation of the method:

[Rianne Margaretha Schouten, Peter Lugtig & Gerko Vink (2018) Generating missing values for simulation purposes: a multivariate amputation procedure, Journal of Statistical Computation and Simulation, 88:15, 2909-2930, DOI: 10.1080/00949655.2018.1491577](https://www.tandfonline.com/doi/full/10.1080/00949655.2018.1491577)

# Multivariate amputation with `ampute`

```{r}
amp <- ampute(data, 
              patterns =  matrix(c(1, 0, 1,
                                   1, 0, 0, 
                                   0, 0, 1), 
                nrow = 3, byrow = TRUE),
              freq = c(0.6, 0.2, 0.2),
              prop = 0.5, 
              mech = "MAR")
md.pattern(amp$amp)
```

# Multivariate amputation with `ampute`

```{r, eval = FALSE}
require(mice)
?ampute
```

`
ampute(data, prop = 0.5, patterns = NULL, freq = NULL, mech = "MAR", weights = NULL, cont = TRUE, type = NULL, odds = NULL, bycases = TRUE, run = TRUE)
`

\vspace{5mm}
Explanation of all the arguments in vignette: 

[https://rianneschouten.github.io/mice_ampute/vignette/ampute.html](https://rianneschouten.github.io/mice_ampute/vignette/ampute.html)

# Missing data mechanisms

Missing Completely At Random (MCAR):

Missingness is not related to any variable

Pr(Income = missing) = 0.5

\vspace{5mm}
Missing At Random (MAR):

Missingness is related to an observed variable

Pr(Income = missing) = Age

\vspace{5mm}
Missing Not At Random (MNAR):

Missingness is related to the missingness itself or to an unobserved variable

Pr(Income = missing) = Income

# Missing data mechanisms: Missing Completely At Random

\begin{figure}
\centering
  \includegraphics[width=110mm]{MCAR.pdf}
\end{figure}

# Missing data mechanisms: Missing At Random

\begin{figure}
\centering
  \includegraphics[width=110mm]{MAR.pdf}
\end{figure}

# Missing data mechanisms: Missing Not At Random

\begin{figure}
\centering
  \includegraphics[width=110mm]{MNAR.pdf}
\end{figure}

# Multivariate amputation with `ampute`

```{r, eval = FALSE}
require(mice)
?ampute
```

`
ampute(data, prop = 0.5, patterns = NULL, freq = NULL, mech = "MAR", weights = NULL, cont = TRUE, type = NULL, odds = NULL, bycases = TRUE, run = TRUE)
`

\vspace{5mm}
Explanation of all the arguments in vignette: 

[https://rianneschouten.github.io/mice_ampute/vignette/ampute.html](https://rianneschouten.github.io/mice_ampute/vignette/ampute.html)

# Contact information

Rianne Schouten: r.m.schouten\@uu.nl

Follow my work: [rianneschouten.github.io](https://rianneschouten.github.io/)

\vspace{1cm}
\begin{figure}
\includegraphics[width=3cm]{uu_logo}\hspace*{2mm}
\includegraphics[width=3cm]{logo_samenveilig}
\end{figure}

# Missingness types

\begin{figure}
\includegraphics{types.pdf}
\end{figure}

# Multivariate Amputation: Weighted sum scores

\begin{itemize}
\item Missing values in multiple variables
\end{itemize}
\begin{equation*}
\begin{blockarray}{cccc}
& Y_1 & Y_2 & Y_3\\
\begin{block}{c[ccc]}
P_1&0&0&1\\
P_2&0&1&1\\
\end{block}
\end{blockarray}
\end{equation*}
\vspace{-5mm}
\begin{itemize}
\item Based on multiple variables \\
$wss_i = w_{1,1} \cdot y_{1i} + w_{1,2} \cdot y_{2i} + w_{1,3} \cdot x_{i} \hspace{2mm} \text{if case $i$ is in pattern 1}$
\end{itemize}
\vspace{2mm}
\begin{equation*}
\begin{blockarray}{cccc}
& Y_1 & Y_2 & Y_3\\
\begin{block}{c[ccc]}
W_1&0&0&1\\
W_2&0&1&5\\
\end{block}
\end{blockarray}
\end{equation*}

# Missing data methods

```{r, include = FALSE}
colors <- c("darkblue", "#b30000")
theme <- list(superpose.symbol = list(col = colors[1:2], pch = 19, cex = 1),
              plot.symbol = list(col = colors[1:2], pch = 1),
              strip.background = list(col = "grey95"))
key <- list(columns = 2, points = list(col = colors[1:2], pch = 19), 
            text = list(c("Observed", "Missing on Income")))

inc_data <- ampute(data, pat = c(0, 1, 1), prop = 0.3, mech = "MNAR")$amp
plot_data <- data[, c(1,3)]
plot_data['R'] <- is.na(inc_data$Income)
plot <- xyplot(Income ~ Age, data = plot_data, 
               groups = R, outer = TRUE, par.settings = theme, 
               ylab = 'Income', xlab = 'Age',
               key = key, ylim = c(-3, 3), xlim = c(-3, 3))
```

```{r, echo = FALSE, fig.width = 10, fig.heigth = 5}
plot
```

# Mean imputation

```{r, include = FALSE}
com_data <- inc_data
com_data[is.na(inc_data$Income), 'Income'] <- 
  mean(inc_data$Income, na.rm = TRUE)
```

```{r, include = FALSE}
key <- list(columns = 2, points = list(col = colors[1:2], pch = 19), 
            text = list(c("Complete", "Imputed")))
com_data['R'] <- is.na(inc_data$Income)
plot <- xyplot(Income ~ Age, data = com_data, 
               groups = R, outer = TRUE, par.settings = theme, 
               ylab = 'Income', xlab = 'Age',
               key = key, ylim = c(-3, 3), xlim = c(-3, 3))
```

```{r, echo = FALSE, fig.width = 10, fig.heigth = 5}
plot
```

# Regression imputation

```{r, include = FALSE}
fit <- lm(Income ~ Age, data = inc_data)
pred <- predict(fit, newdata = ic(inc_data))
com_data <- inc_data
com_data[is.na(inc_data$Income), 'Income'] <- pred
```

```{r, include = FALSE}
com_data['R'] <- is.na(inc_data$Income)
plot <- xyplot(Income ~ Age, data = com_data, 
               groups = R, outer = TRUE, par.settings = theme, 
               ylab = 'Income', xlab = 'Age',
               key = key, ylim = c(-3, 3), xlim = c(-3, 3))
```

```{r, echo = FALSE, fig.width = 10, fig.heigth = 5}
plot
```











