---
title: "Generating missing values with `ampute`"
author: |
  | Rianne Schouten
institute: |
  | 1. Department of Methodology and Statistics, University Utrecht
  | 2. DPA Professionals Data Science Excellence Program
date: "May 18, 2017"
output:
  beamer_presentation: 
    includes:
      in_header: preamble.tex
    theme: default
    colortheme: "beaver"
---

# Overview

```{r, include = FALSE}
require(mice)
require(lattice)
require(MASS)
```

- What is amputation?
- Characteristics of missing data problems
- Stepwise Univariate Amputation
- Multivariate Amputation

```{r, eval = FALSE}
require(mice)
?ampute
```

# What is amputation?

Amputation is the generation of missing values in complete data 

\vspace{5mm}
Evaluation of a missing data methodology:
\begin{itemize}
\item Simulate complete data set
\item Generate missing values
\item Deal with missing data with new method
\item Compare statistical inferences between the original, complete data set and after dealing with the missing values
\end{itemize}

# Characteristics of missing data problems

\begin{align*}
\text{MCAR} &: \text{Missingness is not related to X or Y at all} \\
\text{MAR} &: \text{Missingness is related to X but not to Y} \\
\text{MNAR} &: \text{Missingness is related to Y}\\
\\
\text{MCAR} &: Pr(Y = missing) = 0.5\\
\text{MAR} &: logit(Pr(Y = missing)) = X \\
\text{MNAR} &: logit(Pr(Y = missing)) = Y\\
\end{align*} 

# Characteristics of missing data problems

- Mechanism
- Proportion
- Severity
- Which variables
- Influence of observed data
- Combinations

# Stepwise Univariate Amputation

\begin{columns}

\begin{column}{0.5\textwidth}
\begin{enumerate}
\item $Y_1$
\end{enumerate}
\end{column}

\begin{column}{0.48\textwidth}
\begin{figure}
  \includegraphics[width=100mm, height=60mm]{SUA_1.pdf}
\end{figure}
\end{column}

\end{columns}

# Stepwise Univariate Amputation

\begin{columns}

\begin{column}{0.5\textwidth}
\begin{enumerate}
\item $Y_1$
\item $Y_2$
\end{enumerate}
\end{column}

\begin{column}{0.48\textwidth}
\begin{figure}
  \includegraphics[width=100mm, height=60mm]{SUA_2.pdf}
\end{figure}
\end{column}

\end{columns}

# Stepwise Univariate Amputation

\begin{columns}

\begin{column}{0.5\textwidth}
\begin{enumerate}
\item $Y_1$
\item $Y_2$
\item $\dots$
\item $Y_l$
\end{enumerate}
\end{column}

\begin{column}{0.48\textwidth}
\begin{figure}
  \includegraphics[width=100mm, height=60mm]{SUA.pdf}
\end{figure}
\end{column}

\end{columns}

# Multivariate Amputation

\begin{figure}
\includegraphics[width=115mm, height=63mm]{Scheme_large.pdf}
\end{figure}

# Multivariate Amputation: Missing data patterns

A missing data pattern is a specific combination of variables with missing values and variables without missing values. 

`0`: incomplete variable

`1`: complete variable

# Multivariate Amputation: Missing data patterns

```{r}
data <- mvrnorm(mu = c(5, 5, 10), 
                Sigma = matrix(c(1, 0.5, 0.5, 
                                 0.5, 1, 0.5,
                                 0.5, 0.5, 1), 3, 3), 
                n = 1000)
data <- as.data.frame(data)
names(data) <- c("y1", "y2", "x")
head(data)
```

# Multivariate Amputation: Missing data patterns

```{r}
mypat <- matrix(c(0, 0, 1,
                  0, 1, 1), 
                2, 3, byrow = TRUE)
amp <- ampute(data, patterns = mypat)
inc_data <- amp$amp
```

# Multivariate Amputation: Missing data patterns

```{r}
inc_data[1:12, ]
```

# Multivariate Amputation: Proportion and Frequency

```{r}
amp$prop
amp$freq
```

# Multivariate Amputation: Proportion and Frequency

\begin{figure}
\includegraphics[width=115mm, height=63mm]{Scheme_large.pdf}
\end{figure}

# Multivariate Amputation: Proportion and Frequency

```{r}
myprop <- 0.8
myfreq <- c(0.8, 0.2)
amp <- ampute(data, patterns = mypat, 
              prop = myprop, freq = myfreq)
md.pattern(amp$amp)
```

# Multivariate Amputation

`
ampute(data, prop = 0.5, patterns = NULL, freq = NULL, mech = "MAR", weights = NULL, cont = TRUE, type = NULL, odds = NULL, bycases = TRUE, run = TRUE)
`

```{r}
amp$patterns
```

# Multivariate Amputation: Distribution functions

\begin{figure}
  \includegraphics[width=100mm, height=60mm]{types.pdf}
\end{figure}

```{r}
amp$type
```

# Multivariate Amputation: Weighted sum scores

\begin{itemize}
\item Missing values in multiple variables
\end{itemize}
\begin{equation*}
\begin{blockarray}{cccc}
& Y_1 & Y_2 & X\\
\begin{block}{c[ccc]}
P_1&0&0&1\\
P_2&0&1&1\\
\end{block}
\end{blockarray}
\end{equation*}
\vspace{-5mm}
\begin{itemize}
\item Based on multiple variables \\
$wss_i = w_{1,1} \cdot y_{1i} + w_{1,2} \cdot y_{2i} + w_{1,3} \cdot x_{i} \hspace{2mm} \text{if case $i$ is in pattern 1}$
\end{itemize}
\vspace{2mm}
\begin{equation*}
\begin{blockarray}{cccc}
& Y_1 & Y_2 & X\\
\begin{block}{c[ccc]}
W_1&1&5&0\\
W_2&0&0&1\\
\end{block}
\end{blockarray}
\end{equation*}

# Multivariate Amputation: Weighted sum scores

```{r}
myweights <- matrix(c(1, 5, 0,
                      0, 0, 1), 
                    2, 3, byrow = TRUE)

amp <- ampute(data, patterns =  mypat,
              prop = myprop, freq = myfreq,
              weights = myweights,
              type = c("LEFT", "MID"))
```

# Multivariate Amputation

```{r, echo = FALSE}
bwplot(amp, which.pat = 1, descriptives = FALSE)
```

# Multivariate Amputation

```{r, echo = FALSE}
bwplot(amp, which.pat = 2, descriptives = FALSE)
```

# Multivariate Amputation

- Mechanism
- Proportion
- Severity
- Which variables
- Influence of observed data
- Combinations

`
ampute(data, prop = 0.5, patterns = NULL, freq = NULL, mech = "MAR", weights = NULL, cont = TRUE, type = NULL, odds = NULL, bycases = TRUE, run = TRUE)
`

# Documentation

Rianne Schouten, Peter Lugtig, Jaap Brand, Gerko Vink (2017) Generate missing values with `ampute`, available from: https://github.com/RianneSchouten/Amputation_with_Ampute/tree/master/Manuscript%20article

Rianne Schouten, Peter Lugtig, Gerko Vink (2017) Generating missing values for simulation purposes: A multivariate amputation procedure. Under review. Available from: https://github.com/RianneSchouten/mice/tree/master/vignettes

```{r, eval = FALSE}
require(mice)
?ampute
```

# Additional slides

\begin{figure}
  \includegraphics[width=100mm, height=60mm]{types_shifted.pdf}
\end{figure}

# Additional slides

\begin{figure}
  \includegraphics[width=100mm, height=60mm]{Table.pdf}
\end{figure}

# Characteristics of missing data problems

```{r, include = FALSE}
colors <- c("black", "#fc8d62")
theme <- list(superpose.symbol = list(col = colors, pch = 19, cex = 1),
              plot.symbol = list(col = colors, pch = 1),
              strip.background = list(col = "grey95"))
key <- list(columns = 2, points = list(col = colors, pch = 19), 
            text = list(c("Complete", "Incomplete")))
plotdata <- xyplot(y1 ~ x, data = data, col = "black", ylim = c(2, 8),
                   xlim = c(7, 13), pch = 19)
```

```{r, echo = FALSE, fig.width = 10, fig.heigth = 5}
plotdata
```

# Characteristics of missing data problems: MCAR

```{r}
mcar_data <- ampute(data, pat = c(0, 1, 1), 
                    mech = "MCAR")$amp
```

# Characteristics of missing data problems: MCAR

```{r, include = FALSE}
mis_data <- data
mis_data["R"] <- is.na(mcar_data$y1)
mcar <- xyplot(y1 ~ x, data = mis_data, 
               groups = R, outer = TRUE, par.settings = theme,
               key = key, ylim = c(2, 8), xlim = c(7, 13))
```

```{r, echo = FALSE, fig.width = 10, fig.heigth = 5}
mcar
```

# Characteristics of missing data problems: MAR

```{r}
mar_data <- ampute(data, pat = c(0, 1, 1), 
                   weights = c(0, 0, 1), 
                   mech = "MAR")$amp
```

# Characteristics of missing data problems: MAR

```{r, include = FALSE}
mis_data["R"] <- is.na(mar_data$y1)
mar <- xyplot(y1 ~ x, data = mis_data, 
              groups = R, par.settings = theme,
              key = key, ylim = c(2, 8), xlim = c(7, 13))
```

```{r, echo = FALSE, fig.width = 10, fig.heigth = 5}
mar
```

# Characteristics of missing data problems: MNAR

```{r}
mnar_data <- ampute(data, pat = c(0, 1, 1), 
                    weights = c(1, 0, 0), 
                    mech = "MNAR")$amp
```

# Characteristics of missing data problems: MNAR

```{r, include = FALSE}
mis_data["R"] <- is.na(mnar_data$y1)
mnar <- xyplot(y1 ~ x, data = mis_data, 
               groups = R, par.settings = theme,
               key = key, ylim = c(2, 8), xlim = c(7, 13))
```

```{r, echo = FALSE, fig.width = 10, fig.heigth = 5}
mnar
```

# Documentation

Rianne Schouten, Peter Lugtig, Jaap Brand, Gerko Vink (2017) Generate missing values with `ampute`, available from: https://github.com/RianneSchouten/Amputation_with_Ampute/tree/master/Manuscript%20article

Rianne Schouten, Peter Lugtig, Gerko Vink (2017) Generating missing values for simulation purposes: A multivariate amputation procedure. Under review. Available from: https://github.com/RianneSchouten/mice/tree/master/vignettes

```{r, eval = FALSE}
require(mice)
?ampute
```
